---
title: "Analysis"
author: "Bailey Binando, Lillian Pates"
format: html
editor: visual
---

## Report

## **Stage 3: Analysis or Product**

*Due Thursday, April 30 by class time, on GitHub.*

This stage is more open-ended. Basically, I want you to use skills you’ve learned in this course to do something useful with your data. If you prefer some structure, conduct an Analysis. If you have a different idea you’d like to pursue, it is probably a Product. If you’re unsure whether your idea fits within the scope of this project, ask me about it.

**(Option A) Analysis: Answer a research question.** Start by finalizing a well-defined research question that your data can help you answer. Here are a few examples of good research questions:

1.  (Descriptive) Has wealth inequality increased in the U.S. faster than in Europe since 2000?

2.  (Causal) Does receiving Medicaid coverage reduce the risk of bankruptcy?

3.  (Predictve) Can nighttime satellite imagery be used as a real-time indicator of economic activity?

Then conduct an analysis appropriate to the type of question you are asking.

-   **Descriptive analysis:** Conduct an exploratory analysis of the variables involved in your research question, investigating their distributions and relationships visually. Then use regression modeling to quantify these relationships and conduct statistical inference on them. Write down the regression models you use and interpret your results in words.

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
