---
title: "Stage 2: Visualization"
author: "Bailey Binando, Lillian Pates"
format:
  html: 
    page-layout: full
execute:
  echo: false
  warning: false
  message: false
editor: visual
---

## Report

-   Your topic, motivation, and a brief description of your data and where you got it.

-   An explanation of how you processed your data and why you made the choices you did.

-   The figures you developed (embedded as images).

-   For each figure, explain your intended purpose, why you made the specific visualization choices you did, and how those choices support the figure’s purpose.

```{r package_load}
#| echo: false

# load packages
library(tidyverse)
library(sf) # vector data
library(terra) # rastor data
library(plotly)
library(here)
```

### Figures

```{r data_load}
#| echo: false

# load data

# floodplain shapefiles (2yr, 5yr, 10yr, 25yr, 50yr, 100yr)
floodplain_2yr  <- st_read(here("data", "processed", "floodplain_2yr.gpkg"),  quiet = TRUE)
floodplain_5yr  <- st_read(here("data", "processed", "floodplain_5yr.gpkg"),  quiet = TRUE)
floodplain_10yr <- st_read(here("data", "processed", "floodplain_10yr.gpkg"), quiet = TRUE)
floodplain_25yr <- st_read(here("data", "processed", "floodplain_25yr.gpkg"), quiet = TRUE)
floodplain_50yr <- st_read(here("data", "processed", "floodplain_50yr.gpkg"), quiet = TRUE)
floodplain_100yr <- st_read(here("data", "processed", "floodplain_100yr.gpkg"), quiet = TRUE)

# floodplain asset exposure
floodplain_exposure <- read_csv(here("outputs", "tables", "floodplain_exposure_summary.csv"))

poza_rica <- st_read(here("data", "processed", "poza_rica.gpkg"), quiet = TRUE)

# poza rica building footprint
buildings <- st_read(here("data", "processed", "buildings_clipped.gpkg"), quiet = TRUE)

# poza rica total assets
poza_rica_assets <- read_csv(here("outputs", "tables", "poza_rica_asset_summary.csv"))

# building exposure/vulnerability
exposure_vulnerability <- st_read(here("data", "processed", "building_exposure_vulnerability.gpkg"), quiet = TRUE)

# dem
dem <- rast(here("data", "processed", "dem_pozarica.tif"))

# river
river <- st_read(here("data", "processed", "river_clipped.gpkg"), quiet = TRUE)

# flood timeseries
flood_flow <- read_csv(here("data", "processed", "flood_timeseries_clean.csv"))

```

#### Spatial Analysis

```{r}
#| label: fig1_exposed_vars

# create variables to be used in hover text labels in fig1

# total assets
total_buildings <- poza_rica_assets$total_buildings
total_building_area <- round(poza_rica_assets$total_building_area, 0)

# 100yr floodplain
ratio_buildings_fp100 <- floodplain_exposure|> # building ratio in fp
  filter(floodplain == "fp100") |>
  pull(buildings_exposed_ratio)

pct_buildings_fp100 <- round(ratio_buildings_fp100 * 100, 2) # convert to percent

ratio_area_fp100 <- floodplain_exposure|> # building area ratio in fp
  filter(floodplain == "fp100") |>
  pull(building_area_exposed_ratio)

pct_area_fp100 <- round(ratio_area_fp100 * 100, 2) # convert to percent

# 50yr floodplain
ratio_buildings_fp50 <- floodplain_exposure|>
  filter(floodplain == "fp50") |>
  pull(buildings_exposed_ratio)

pct_buildings_fp50 <- round(ratio_buildings_fp50 * 100, 2)

ratio_area_fp50 <- floodplain_exposure|>
  filter(floodplain == "fp50") |>
  pull(building_area_exposed_ratio)

pct_area_fp50 <- round(ratio_area_fp50 * 100, 2)

# 25yr floodplain
ratio_buildings_fp25 <- floodplain_exposure|>
  filter(floodplain == "fp25") |>
  pull(buildings_exposed_ratio)

pct_buildings_fp25 <- round(ratio_buildings_fp25 * 100, 2)

ratio_area_fp25 <- floodplain_exposure|>
  filter(floodplain == "fp25") |>
  pull(building_area_exposed_ratio)

pct_area_fp25 <- round(ratio_area_fp25 * 100, 2)

# 10yr floodplain
ratio_buildings_fp10 <- floodplain_exposure|>
  filter(floodplain == "fp10") |>
  pull(buildings_exposed_ratio)

pct_buildings_fp10 <- round(ratio_buildings_fp10 * 100, 2)

ratio_area_fp10 <- floodplain_exposure|>
  filter(floodplain == "fp10") |>
  pull(building_area_exposed_ratio)

pct_area_fp10 <- round(ratio_area_fp10 * 100, 2)

# 5yr floodplain
ratio_buildings_fp5 <- floodplain_exposure|>
  filter(floodplain == "fp5") |>
  pull(buildings_exposed_ratio)

pct_buildings_fp5 <- round(ratio_buildings_fp5 * 100, 2)

ratio_area_fp5 <- floodplain_exposure|>
  filter(floodplain == "fp5") |>
  pull(building_area_exposed_ratio)

pct_area_fp5 <- round(ratio_area_fp5 * 100, 2)

# 2yr floodplain
ratio_buildings_fp2 <- floodplain_exposure|>
  filter(floodplain == "fp2") |>
  pull(buildings_exposed_ratio)

pct_buildings_fp2 <- round(ratio_buildings_fp2 * 100, 2)

ratio_area_fp2 <- floodplain_exposure|>
  filter(floodplain == "fp2") |>
  pull(building_area_exposed_ratio)

pct_area_fp2 <- round(ratio_area_fp2 * 100, 2)

```

```{r}
#| label: fig-floodplain_exposure_map
#| out-width: "100%"
#| fig-height: 6
#| fig-cap: "Descriptive spatial analysis of asset exposure across varying floodplain extents. Darker shades indicate more frequently flooded zones in Poza Rica, Veracruz, México. <br>Data Sources: Microsoft Building Footprints, HAND Floodplain, OCHA, Gobjerno de México "

# create color palette for floodplains (darker for area with more frequent flooding)
cols <- c(
  "Floodplain (100yr)" = "#9ecae1",
  "Floodplain (50yr)"  = "#6baed6",
  "Floodplain (25yr)"  = "#4292c6",
  "Floodplain (10yr)"  = "#2171b5",
  "Floodplain (5yr)"   = "#08519c",
  "Floodplain (2yr)"   = "#08306b"
)

# create spatial map of floodplains, buildings, and river
map <- ggplot() +
  geom_sf(data = poza_rica, 
          aes(text = paste0("Municipal Boundary", # hover text of total assets
                            "<br>Population: 189,457 (2020 Census)",
                            "<br>Buildings: ", total_buildings,
                            "<br>Building Area: ", total_building_area, " m²")), 
          fill = "#f7f3ea", color = "grey70", linewidth = 0.2) +
  
  geom_sf(data = floodplain_100yr, 
          aes(fill = "Floodplain (100yr)",
              text = paste0("100-Year Floodplain", # hover text of exposed assets
                            "<br>Buildings Exposed: ", pct_buildings_fp100, "%",
                            "<br>Building Area Exposed: ", pct_area_fp100, "%")), 
          alpha = 0.3, color = NA) +
  
  geom_sf(data = floodplain_50yr, 
          aes(fill = "Floodplain (50yr)",
              text = paste0("50-Year Floodplain", # hover text of exposed assets
                            "<br>Buildings Exposed: ", pct_buildings_fp50, "%",
                            "<br>Building Area Exposed: ", pct_area_fp50, "%")),
          alpha = 0.24, color = NA) +
  
  geom_sf(data = floodplain_25yr, 
          aes(fill = "Floodplain (25yr)",
              text = paste0("25-Year Floodplain", # hover text of exposed assets
                            "<br>Buildings Exposed: ", pct_buildings_fp25, "%",
                            "<br>Building Area Exposed: ", pct_area_fp25, "%")),
          alpha = 0.26, color = NA) +
  
  geom_sf(data = floodplain_10yr, 
          aes(fill = "Floodplain (10yr)",
              text = paste0("10-Year Floodplain", # hover text of exposed assets
                            "<br>Buildings Exposed: ", pct_buildings_fp10, "%",
                            "<br>Building Area Exposed: ", pct_area_fp10, "%")),
          alpha = 0.28, color = NA) +
  
  geom_sf(data = floodplain_5yr, 
          aes(fill = "Floodplain (5yr)",
              text = paste0("5-Year Floodplain", # hover text of exposed assets
                            "<br>Buildings Exposed: ", pct_buildings_fp5, "%",
                            "<br>Building Area Exposed: ", pct_area_fp5, "%")),
          alpha = 0.30, color = NA) +
  
  geom_sf(data = floodplain_2yr, 
          aes(fill = "Floodplain (2yr)",
              text = paste0("2-Year Floodplain", # hover text of exposed assets
                            "<br>Buildings Exposed: ", pct_buildings_fp2, "%",
                            "<br>Building Area Exposed: ", pct_area_fp2, "%")),
          alpha = 0.32, color = NA) +
  
  geom_sf(data = buildings, color = "grey40", alpha = 0.7, size = 0.06) +
  geom_sf(data = river, color = "#08306b",
          aes(text = "Río Cazones"),
          linewidth = 0.65) +
  labs(
    title = "Building Exposure Across Floodplain Return Periods",
    fill = "Flood Extent"
  ) +
  scale_fill_manual(
    values = cols
  ) +
  theme_bw() +
  theme(panel.grid = element_blank(), # remove grid
        plot.title = element_text(hjust = 0.5) # center title
  )

# make iteractive w/ plotly
map_plotly <- ggplotly(map, tooltip = "text") # add hover label


# use plotly to set range of x and y axis
# xlim/ylim/coord_sf did not work (issues w/ ggplot --> plotly)
fig1 <- layout(map_plotly,
  xaxis = list(range = c(660670, 666857)),
  yaxis = list(range = c(2266165, 2278400))
)

fig1


```

words

```{r}
#| label: fig-elevation_map
#| out-width: "100%"
#| fig-height: 6
#| fig-cap: "Digital elevation map (DEM) of Poza Rica, Veracruz, México. 
#| <br>Data Sources: NASA Earthdata, OCHA"

# convert dem into a dataframe
dem_df <- as.data.frame(dem, xy = TRUE, na.rm = TRUE)

dem_df <- dem_df |>
  rename(elevation = n20_w100_1arc_v3) # rename var

# plot elevation map
elevation_map <- ggplot() +
  geom_raster(data = dem_df, aes(x = x, y = y, fill = elevation,
      text = paste0("Elevation: ", round(elevation, 2), " m"))) + 
  coord_equal() + # prevent distortion
  geom_sf(data = poza_rica, color = "black", linewidth = 0.2, fill = NA) + # add municipality outline
  scale_fill_viridis_c(name = "Elevation (m)") +
  labs(
    title = "Elevation Map"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(), # remove gridlines
    plot.title = element_text(hjust = 0.5), # center title
    axis.title = element_blank() # remove axis labels
  )

# make interactive
fig2 <- ggplotly(elevation_map, tooltip = "text") # add hover label

fig2
```

words

```{r}
#| label: fig-histograms_of_key_vars
#| fig-width: 12
#| fig-height: 6
#| fig-cap: "Distribution of Exposure & Vulnerability Indicators 
#| <br>Data Sources: Microsoft Building Footprints, NASA Earthdata" 

exposure_vulnerability |>
  st_drop_geometry() |>
  select(river_distance, building_elevation, building_slope) |>
  pivot_longer(everything()) |>
  mutate(name = recode(name,
    "river_distance"    = "Distance to River (m)",
    "building_elevation" = "Building Elevation (m)",
    "building_slope"    = "Building Slope (°)"
  )) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "#0A7398", color = "white") +
  facet_wrap(~name, scales = "free") +
  labs(title= "Histograms",  x = NULL, y = "Count") +
  theme_bw()+
  theme(
    panel.grid = element_blank(), # remove gridlines
    plot.title = element_text(hjust = 0.5), # center title
  )
```

words

```{r}
#| label: fig-elevation_distance
#| fig-height: 16
#| fig-width: 14
#| fig-cap: "Hexbin plot of Poza Rica mean exposure ratio across building distance from Río Cazones and elevation. Mean exposure is the ratio of building surface area exposed to flooding. The majority of building exposure lies within an elevation ~80m. <br>Data Sources: Microsoft Building Footprints, NASA Earthdata, OCHA, HAND Floodplain"

elevation_distance <- exposure_vulnerability |>
  st_drop_geometry() |>
  select(building_elevation, river_distance, exposure_2yr, exposure_5yr, exposure_10yr,
         exposure_25yr, exposure_50yr, exposure_100yr) |>
  pivot_longer(
    cols = c(exposure_2yr, exposure_5yr, exposure_10yr, exposure_25yr, 
             exposure_50yr, exposure_100yr),
    names_to = "floodplain",
    values_to = "exposure"
  ) |>
  mutate(
    floodplain = factor(floodplain,
                        levels = c("exposure_2yr", "exposure_5yr", "exposure_10yr", 
                                   "exposure_25yr", "exposure_50yr", "exposure_100yr"))
  )

hexbin <- ggplot(elevation_distance, aes(x = building_elevation, y = river_distance)) +
  stat_summary_hex(
    aes(z = exposure, text = after_stat(paste0("Mean Exposure: ", round(value, 2)))),
    fun = mean,
    bins = 20
  ) +
  scale_fill_viridis_c(name = "Mean Exposure Ratio", option = "plasma") +
  labs(
    x = "Elevation (m)",
    y = "Distance to River (m)",
    title = "Effect of Building Distance and Elevation on Flood Exposure"
  ) +
  # adds a line
  annotate("segment", x = 80, y = 47.614, xend = 80, 
           yend = 8200, color = "grey60", linewidth = 1, linetype = "dashed") +
  # adds text annotation of street name
  annotate("text", x = 80, y = 8200, label = "Elevation Threshold: ~80m", size = 3) +
  theme_bw() +
  theme(
    panel.grid = element_blank(), # remove gridlines
    plot.title = element_text(hjust = 0.5), # center title
  ) +
  facet_wrap(~ floodplain,
             labeller = as_labeller(c("exposure_2yr" = "2-Year Flood",
                                      "exposure_5yr" = "5-Year Flood",
                                      "exposure_10yr" = "10-Year Flood ", 
                                      "exposure_25yr" = "25-Year Flood", 
                                      "exposure_50yr" = "50-Year Flood", 
                                      "exposure_100yr" = "100-Year Flood")))

fig4 <- ggplotly(hexbin, tooltip = "text")

fig4

```

Sources:

<https://carpentry.library.ucsb.edu/2022-08-09-ucsb-r-geospatial/02-raster-plot/>
